---
title: "Perry_analysis"
author: "Madelyn Willis"
date: "`r Sys.Date()`"
output: html_document
---

## Unloading Necessary Packages
```{r Packages, include=FALSE}
# Load required libraries
library(sf)
library(tidyverse)
library(terra)
library(caret)
library(randomForest)
library(ggplot2)
```

## Data Preparation
```{r Data Preparation}
# Load spatial and tabular data
soil_points <- st_read("../Local Data- DO NOT PUSH/surface_points_perry_corrected.shp") %>%
  pivot_longer(cols = c("CID", "SPID"), names_to = "type", values_to = "ID")
soil_all <- read.csv("../Local Data- DO NOT PUSH/Perry_all_samples_all(in).csv")
soil_points <- merge(soil_points, soil_all, by = "ID")

# Load raster covariates
covariates <- c(
  rast("../Local Data- DO NOT PUSH/PCA_SR.tif"),
  rast("../Local Data- DO NOT PUSH/clipped_stack.tif"),
  rast("../Local Data- DO NOT PUSH/NDVI.tif")
)
names(covariates) <- make.names(names(covariates))

# Extract raster values at soil sample points
soil_points_sp <- vect(soil_points)
soil_points_extracted <- terra::extract(covariates, soil_points_sp, bind = TRUE, xy = TRUE) %>%
  as.data.frame()

# Filter relevant columns
soil_vars <- c("LBC.1", "pH.2", "Ca", "K", "Mg", "Mn", "P", "Zn")
soil_data <- soil_points_extracted[, c(soil_vars, names(covariates))]
```

## Train-Test Split
```{r Train-Test Split}
# Perform an 80/20 split of the entire dataset
set.seed(626)  # For reproducibility
split <- caret::createDataPartition(1:nrow(soil_data), p = 0.8, list = FALSE)
train_data <- soil_data[split, ]
test_data <- soil_data[-split, ]
```

## Random Forest and Prediction
```{r Random Forest and Prediction}
run_rf <- function(response, train_data, covariates, output_path) {
  # Ensure the response variable exists in train_data
  if (!response %in% colnames(train_data)) {
    stop(paste("The response variable", response, "is not found in the training data."))
  }
  
  # Exclude nutrient variables from predictors
  predictors <- setdiff(colnames(train_data), soil_vars)
  
  # Train random forest model
  train_control <- trainControl(method = "cv", number = 5)
  rf_model <- train(reformulate(termlabels = predictors, response = response), 
                    data = train_data, method = "rf", trControl = train_control)
  
  # Predict raster values
  prediction <- terra::predict(covariates, rf_model, na.rm = TRUE)
  terra::writeRaster(prediction, output_path, overwrite = TRUE)
  return(prediction)
}

# Loop through all nutrients and train models
for (var in soil_vars) {
  rf_output_path <- paste0("../Local Data- DO NOT PUSH/rf nutrient rasters/", var, "_rf.tif")
  print(paste("Processing nutrient:", var))
  prediction <- run_rf(var, train_data, covariates, rf_output_path)
  plot(prediction, main = paste(var, "Random Forest Prediction"))
}
```

## Kriging for Single Points
```{r Kriging Single Points}
# Nutrients and their variogram models
variogram_models <- c(LBC.1 = "Sph", pH.2 = "Exp", Ca = "Sph", 
                      K = "Lin", Mg = "Sph", Mn = "Nug", 
                      P = "Lin", Zn = "Nug")

# Loop through nutrients and perform kriging
for (var in names(variogram_models)) {
  tryCatch({
    formula <- as.formula(paste(var, "~ 1"))
    krig_output_path <- paste0("../Local Data- DO NOT PUSH/krig rasters/single point/", var, "_krig_SP.tif")
    print(paste("Processing kriging for single point nutrient:", var))
    kriged <- run_kriging(formula, soil_points_single, grid, variogram_models[[var]], krig_output_path)
    plot(kriged, main = paste(var, "Kriging - Single Points"))
  }, error = function(e) {
    print(paste("Error processing kriging for single point nutrient:", var, " - ", e$message))
  })
}
```

## Kriging for Composite Points
```{r Kriging Composite Points}
# Loop through nutrients and perform kriging
for (var in names(variogram_models)) {
  tryCatch({
    formula <- as.formula(paste(var, "~ 1"))
    krig_output_path <- paste0("../Local Data- DO NOT PUSH/krig rasters/composite/", var, "_krig_C.tif")
    print(paste("Processing kriging for composite point nutrient:", var))
    kriged <- run_kriging(formula, soil_points_composite, grid, variogram_models[[var]], krig_output_path)
    plot(kriged, main = paste(var, "Kriging - Composite Points"))
  }, error = function(e) {
    print(paste("Error processing kriging for composite point nutrient:", var, " - ", e$message))
  })
}
```

---
---

### Key Features of the Updated Workflow

1. **Single Train-Test Split**:
   - A single 80/20 split is done on the dataset using `caret::createDataPartition`.
   - The same training set (`train_data`) is used to train all nutrient models.

2. **Random Forest Training and Prediction**:
   - A function `run_rf` handles training and prediction for each nutrient.
   - The corresponding nutrient raster is saved to disk using `terra::writeRaster`.

3. **Streamlined Loop**:
   - A single loop iterates through all nutrient variables (`soil_vars`).
   - Each nutrient is trained and predicted independently using the same `train_data`.

### Output
This workflow ensures that:
- Every nutrient is modeled and predicted independently.
- A single train-test split is used across all nutrients.
- Predicted rasters for each nutrient are saved as GeoTIFF files.

Let me know if you need further adjustments!